\documentclass[12pt]{report}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{enumitem}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{multirow}
\usepackage{longtable}
\usepackage{subcaption}
\usepackage{float}
\usepackage[normalem]{ulem}
\graphicspath{ {images/} }
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{15pt}
\lhead{\rmfamily \small \nouppercase \leftmark}
\rfoot{\thepage}
\lfoot{\rmfamily \small \nouppercase \rightmark}
\renewcommand{\headrulewidth}{2pt}
\renewcommand{\footrulewidth}{1pt}
\DeclareMathOperator*{\rint}{\ThisStyle{\hstretch{1.3}{\rotatebox{18}{$\SavedStyle\!\int\!$}}}}
% \usepackage{biblatex}
\usepackage[backend=biber,sorting=none,style=phys,%
articletitle=false,biblabel=brackets,%
chaptertitle=false,pageranges=false]{biblatex}
\addbibresource{thesisbib.bib}
\usepackage[usenames,dvipsnames,x11names]{xcolor}
\usepackage{tikz}
\usetikzlibrary{arrows,shapes,decorations.pathmorphing,positioning,chains}
% Set up a few colours
\colorlet{lcfree}{Green3}
\colorlet{lcnorm}{Blue3}
\colorlet{lccong}{Red3}
\addbibresource{./thesisbib.bib}
\pgfdeclarelayer{marx}
\pgfsetlayers{main,marx}
% A macro for marking coordinates (specific to the coordinate naming
% scheme used here). Swap the following 2 definitions to deactivate
% marks.
\providecommand{\cmark}[2][]{%
  \begin{pgfonlayer}{marx}
    \node [nmark,yshift=1em] at (c#2#1) {#2};
  \end{pgfonlayer}{marx}
  } 
\providecommand{\cmark}[2][]{\relax} 
\newcommand{\proton}[1]{%
	\shade[ball color=red] (#1) circle (.25);
}

%\neutron{xposition,yposition}
\newcommand{\neutron}[1]{%
	\shade[ball color=green] (#1) circle (.25);
}
\newcommand{\pion}[1]{%
	\shade[ball color=blue] (#1) circle (.15);
}

%\electron{xwidth,ywidth,rotation angle}
\newcommand{\electron}[3]{%
	\draw[rotate = #3](0,0) ellipse (#1 and #2)[color=blue];
	\shade[ball color=Gold2] (0,#2)[rotate=#3] circle (.1);
}
\newcommand{\nucleust}{%
	\neutron{7.2,-1.3}
	\proton{7,-1}
	\neutron{7.3,-1.2}
	\proton{6.8,-1.1}
	\neutron{6.9,-1.3}
	\proton{7.2,-1.15}
	\neutron{6.95,-1.12}
}

\newcommand{\nucleus}{%
	\neutron{0.2,0.3}
	\proton{0,0}
	\neutron{0.3,0.2}
	\proton{-0.2,0.1}
	\neutron{-0.1,0.3}
	\proton{0.2,-0.15}
	\neutron{-0.05,-0.12}
	\proton{0.17,0.21}
	\neutron{-0.21,0.3}
	\proton{0.27,0.21}
	\neutron{0.21,0.3}
	\proton{0.17,0.31}
}



\begin{document}
\input{chapters/titlepage}
\tableofcontents
\chapter*{Abstract}

Studies of the selected problems in understanding of nuclear spallation physics are undertaken. Among them are the description of the initial phase of proton - target nucleus collision and proceeding of the intranuclear cascade, isotropic emission of nuclear fragments of very broad mass spectrum, contribution of the nonequlibrium and equilibrium processes to the total production cross-section and its dependence on the isotope isospin, 
variation of the total production cross-section related to the number and ratio of protons and neutrons of the  emitted particle. 
New experimental distributions of the double differential production cross-sections ($d^2\sigma/d\Omega dE$) for $p$,  $d$,  $t$, $\pi^{+}$ and $\pi^{-}$ in the $p+Nb$ reaction 
at 3.5 GeV proton beam energy are provided. They have been measured by High Acceptance DiElectron Spectrometer (HADES) experiment. The measured energy range of hydrogen isotopes 
are significantly extended in comparison to the experimental data available in the literature. The new experimental spectra are confronted with the theoretical prediction of three models of intranuclear cascade - GiBUU, UrQMD, INCL++. Discrepancies
of the experimental and theoretical distributions are noticed  and discussed. Experimental data derived from the literature are used for examination of nuclear mechanisms assigned to the nuclear deexcitation phenomena. The comparative analysis of the four model prediction (ABLA07, GEMINI++, GEM2, SMM) of the selected observables relevant to the isotropic nuclear fragment emission is performed. Disagreement of the model prediction and the experimental data is observed. 


%The aim of the present work was to assess the model capabilities in describing the proton induced spallation reactions. To understand the reaction mechanism, different sets of observables must be investigated, inclusive as well as exclusive. Based on these criteria, the eforts were done to survey the scientific literature for the selection of representative  data sets which its the need. The selected data were rich in terms of production of various ejectiles: neutrons, light charged particles (LCP: p, d, t, 3He, 4He), intermediate mass fragments, i.e., the particles with atomic mass number (LCP, IMF , fission fragments), and target-like heavy residua. Several atomic nuclei from Al up to Pb were selected as representative for all the targets. The proton beam covered the broad range of energies from 180 MeV to 3000 MeV. The spallation reaction was treated as a two stage process. In the first stage, the incident proton initiates the cascade of binary collisions with target nucleons leaving
%behind an excited remnant. The second stage consists in the decay of this excited remnant nucleus. The selection of best models to describe each of these two stages was done on
%the basis of previous benchmark eforts where INCL4.5 model was found to be the best to describe the first stage of the reaction. Therefore the newest version of this model -
%INCL4.6 was used in the present study. Four theoretical models different in approach  to the reaction mechanism were chosen to realize the description of the second stage: ABLA07, GEMINI++, SMM, GEM2. Qualitative as well as quantitative comparisons of model calculations with experimental data were undertaken. To judge the quality of models the agreement in magnitude of different observables with model predictions as
%well as reproduction of the shape of the mass, angle and energy distributions of the cross-sections were taken into account. Various deviation factors were used for providing ranking and validation of the spallation models. The statistical properties of the test factors
 
\chapter*{Dedication}
This work is dedicated to my late mother. Also to my father, my younger brothers and my late supervisor Prof. dr. hab. Zbigniew Rudy.
 
%\chapter*{Declaration}
%I declare that..
 
\chapter*{Acknowledgements}
First of all, I would like to thank my supervisor and co-supervisor, Prof. dr. hab. Krzysztof Pysz and Dr. Shushil Sharma for guidance and dedicated involvement in every step throughout the process. I also wanted to thank Prof. dr. hab. Boguław Kamys for his guidance, motivation and support \par
I would thank Prof. dr. hab Piotr Salabura, Dr. hab. Witold Przygoda, Dr. Lorenz Manuel, Dr. Izabela Ciepal, Rafal Lalik and all the people from HADES collaboration for guidance and support.\par
I also like to thanks all my colleagues from our department of physics for their motivation and support.\par
Last but not least, I would like to thank Anjali Aggarwal. She made me smile, motivated and supported me all those years without her. I wouldn't complete this work.
\chapter{Introduction}\label{chapter1}
\input{chapters/introduction}
 
\chapter{Contemporary theoretical models for spallation reactions}\label{chapter:2}
\input{chapters/chapter02}
\chapter{Mechanisms of proton-nuclear target interaction - studies with HADES spectrometer}\label{chapter:3} %– anisotropic production of pions and H
% isotopes}
\input{chapters/chapter03}
\chapter{Results of HADES experiment}\label{chapter:4}
\input{chapters/chapter04}
  \chapter{Mechanisms of isotropic emission of IMF and heavy target remnants}\label{chapter:5}
  \input{chapters/chapter05}
\chapter{The yields of non-equilibrium and equilibrium processes in spallation reactions (IMF)}\label{chapter:6}
 \input{chapters/chapter06.tex}
\chapter{Odd-even staggering of total production cross-section in spallation reactions}\label{chapter:7}
 \input{chapters/chapter07.tex}
\chapter{Summary}
\input{chapters/conclusion}
\printbibliography
\end{document}
% \appendix
% \chapter{Appendix Title}
% \input{chapters/appendix}